---
title: "Story - 1 : Infrastructure Investment & Jobs Act Funding Allocation"
author: 'Daria Dubovskaia'
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: united
    code_folding: hide
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
# chunks
knitr::opts_chunk$set(eval=TRUE, message=FALSE, warning=FALSE, fig.height=5, fig.align='center')

# libraries
library(viridis)
library(tidyverse)
library(caret)
library(ggplot2)
library(openxlsx)
library(kableExtra)
library(summarytools)
library(ggpubr)


# random seed
set.seed(42)
```

```{r common functions}
#' nice_table
#' 
#' @param df
#' @param fw
nice_table <- function(df, cap=NULL, cols=NULL, dig=3, fw=F){
  if (is.null(cols)) {c <- colnames(df)} else {c <- cols}
  table <- df %>% 
    kable(caption=cap, col.names=c, digits=dig) %>% 
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      html_font = 'monospace',
      full_width = fw)
  return(table)
}
```


## Overview
The Infrastructure Investment and Jobs Act is a significant commitment by the federal government to improving the nation's infrastructure. This analysis seeks to assess the equity of the Act's funding allocation across states and territories, as well as whether there is any evidence of political bias in the distribution of funds. Understanding how these funds are allocated is critical for ensuring that the Act achieves its stated goal of equitable infrastructure development.

## 1. Data Preparation

### 1.1 Summary Statistics

**Funding Data**

The funding data contained **57** observations of 2 variables. It included columns for each state, territory, or tribal nation, as well as total funding in billions. The data source: [Github repo](https://github.com/ex-pr/DATA_608/blob/main/IIJA%20FUNDING%20AS%20OF%20MARCH%202023(1).xlsx)

**Population Data**

The population data had a complex structure, with row headers and numerous unnamed columns. The relevant data appeared to be organized into columns with headers such as "April 1, 2020 Population Estimates Base" and "Population Estimate (as of July 1)" for various years. As a results of data preparation step, this dataset should have contained **57** observations of 2 variables. It included columns for each state, territory, or tribal nation, as well as total population for 2020 year. The data source: [United States Census Bureau](https://www.census.gov/data/datasets/time-series/demo/popest/2020s-state-total.html), [List of U.S. states and territories by population](https://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_population), [Native Americans in the United States](https://en.wikipedia.org/wiki/Native_Americans_in_the_United_States)


**Election Results Data**

The election results data included columns for the state, final call (indicating which party won), Democratic and Republican votes, other votes, percentages for each, and margin of victory. There were additional columns for state IDs, electoral votes, and other metadata.  As a results of data preparation step, this dataset should have contained **51** observations of 4 variables. The data source: [The Cook Political Report with Amy Walter](https://www.cookpolitical.com/2020-national-popular-vote-tracker)

All three datasets had no missing values for their key columns, indicating that the data was complete in terms of the presence of values.


```{r data_load}
# Load data from Github
df1 <- read.csv("https://raw.githubusercontent.com/ex-pr/DATA_608/main/Popular%20vote%20backend%20-%20Sheet1.csv")

df2 <- read.xlsx("https://github.com/ex-pr/DATA_608/blob/main/NST-EST2023-CHG.xlsx?raw=true", 1)

df3 <- read.xlsx("https://github.com/ex-pr/DATA_608/blob/main/IIJA%20FUNDING%20AS%20OF%20MARCH%202023(1).xlsx?raw=true", 1)
```

```{r copy_data}
elections <- df1
population <- df2
funding <- df3
```


```{r summary_data}
#show the data in tables
elections %>% 
  nice_table(cap='The 2020 Presidential election') %>% 
  scroll_box(width='100%', height = "400px")

population %>% 
  nice_table(cap='State Population Totals and Components of Change: 2020-2023') %>% 
  scroll_box(width='100%',  height = "400px")

funding %>% 
  nice_table(cap='Allocation of the Infrastructure Investment and Jobs Act, 2023') %>% 
  scroll_box(width='100%',  height = "400px")
```

### 1.2 Column change

The funding dataset's column names have been changed to more descriptive and consistent titles, making the data easier to access and understand.To ensure consistency and accuracy in state naming, "DELEWARE" was corrected to "Delaware", as were "District Of Columbia" to "District of Columbia" and "Us Virgin Islands" to "US Virgin Islands". State names were converted from uppercase to title case (first letter uppercase, rest lowercase) to ensure consistency across datasets.

```{r clear_funding}
#change column names
funding <- funding %>% 
  rename("state" = "State,.Teritory.or.Tribal.Nation", "total_billions" = "Total.(Billions)")

#change DELEWARE to Delaware
funding[funding=="DELEWARE"] <- "Delaware"

#change state names from uppercase to lower
funding$state <- str_to_title(funding$state)

#change DELEWARE to Delaware
funding[funding=="District Of Columbia"] <- "District of Columbia"
funding[funding=="Us Virgin Islands"] <- "US Virgin Islands"
```

For the population dataset, only relevant columns, specifically those containing state names and population figures, were kept and renamed for clarity. The index column was reset to ensure that rows were numbered correctly and sequentially after selecting a specific range of rows.Dots in state names were removed to ensure consistency with the naming convention used in other datasets.
A new dataframe was created to include U.S. territories and tribal communities that were not previously included in the population dataset, ensuring complete coverage of all regions impacted by the funding.
The population figures were converted to numeric data types, allowing for mathematical operations and analyses.

```{r clear_population}
#choose columns with states only
population <- population[10:61,1:2] %>% 
  rename("state" = 1, "total_population" = 2)

#reset index column
rownames(population) <- NULL 

#remove dots from the state names
population$state <- gsub("\\.", "", population$state) 

#new df with us territories and population
df <- data.frame("state" = c("Guam", "US Virgin Islands", "American Samoa", "Northern Mariana Islands", "Tribal Communities"),
                  "total_population" = c(153836, 87146, 49710, 47329, 9666058))

#add us territories to the data
population <- rbind(population,df)

#transform total_population column to numeric
population$total_population <- as.numeric(population$total_population)
```

For the elections dataset, to ensure consistency, it was filtered to remove unnecessary rows and columns, focusing only on relevant states and removing districts within states except for the District of Columbia. Results for Democratic or Republican wins were simplified to "Biden" or "Trump" to facilitate analysis. Vote counts were originally in string format with commas but were converted to numeric values to allow for quantitative comparisons.

```{r clear_elections}
#choose columns with states only
elections <- elections[5:61,] %>%
  filter(!row_number() %in% c(16))

#remove districts for Nebraska, Maine but keep District of Columbia
columbia <- elections %>% 
  filter(str_detect(state, "District of Columbia"))

elections <- elections[!grepl("District", elections$state), ]

elections <- rbind(elections, columbia)

#choose columns with results for democrats or republicans only
elections <- elections[,c(1, 2, 4, 5)]

#substitute D, R for results with democrats win (1) or not (0)
elections <- elections %>% mutate(win = factor(ifelse(called == "D","Biden","Trump"))) %>%
  select(-called)

#transform to numeric values columns with votes
elections$dem_votes <- as.numeric(gsub(",", "", elections$dem_votes))
elections$rep_votes <- as.numeric(gsub(",", "", elections$rep_votes))
```


### 1.3 Merge data. New features
We also calculated the funding per capita for each state and territory by dividing the total funding by the population. This helped us assess the equity of funding allocation relative to population size. Furthermore, to improve clarity in subsequent visualizations, state names were substituted with their respective two-letter abbreviations. The datasets had been successfully merged, and the funding per capita for each state and territory had been calculated. The resulting dataset contained information such as state names, total funding in billions of dollars, population figures, an indicator of whether Biden or Trump won each state in the 2020 election, and calculated funding per capita. This extensive dataset served as the foundation for subsequent analyses, visualizations, and interpretations aimed at gaining insight into the Infrastructure Investment and Jobs Act's funding allocation.


```{r merge_data}
#merge funding and population data
first_df <- merge(x = funding, y = population, by = "state", all = TRUE)

#add new feature
first_df <- first_df %>%
  mutate(funding_per_capita = total_billions*1000000000 / total_population)

#state names to abbriviations
first_df$abbriv <- state.abb[match(first_df$state,state.name)]
first_df <- first_df %>%
     mutate(abbriv=replace(abbriv, state=="American Samoa", "AS")) %>%
     mutate(abbriv=replace(abbriv, state=="District of Columbia", "DC")) %>%
     mutate(abbriv=replace(abbriv, state=="Guam", "GU")) %>% 
  	mutate(abbriv=replace(abbriv, state=="Northern Mariana Islands", "MP")) %>% 
     mutate(abbriv=replace(abbriv, state=="Puerto Rico", "PR")) %>%
     mutate(abbriv=replace(abbriv, state=="Tribal Communities", "TT")) %>%     
     mutate(abbriv=replace(abbriv, state=="US Virgin Islands", "VI")) 

#merge funding, population and election data
second_df <- merge(x = first_df, y = elections, by = "state",  all.y = TRUE)

second_df %>% 
  nice_table(cap='Merged dataset') %>% 
  scroll_box(width='100%', height = "400px")
```



## 2. Data analysis

For our visual analysis, we used bar plots to represent funding per capita because they clearly show comparative differences between states. Scatter plots were used to visually represent the relationship between funding per capita and population size, taking advantage of their ability to display correlations between two quantitative variables. These plots were chosen to effectively communicate funding distribution and disparities.


### 2.1 Question 1: Equitable allocation based on population
 
Our detailed examination of each state's funding per capita in relation to population size revealed significant disparities. The visualization of funding per capita by state/territory reveals the uneven distribution of funds, challenging the notion that allocation was solely determined by population.

The funding per capita by state/territory visualization clearly depicted the distribution of funds relative to the population. We saw that funding per capita varied by state, with Alaska, Wyoming, and North Dakota receiving the most. This could imply that allocation was influenced by other factors such as geographical location, infrastructure requirements, or even the cost of living and construction in these areas. The next bar chart showed the populations of each state and territory. The stark contrast in population sizes across the United States was immediately noticeable, with states such as California and Texas dwarfing others. Notably, smaller states and territories, despite having fewer people, sometimes received more funding per capita, implying a complex allocation mechanism that takes into account factors other than sheer population.

For example, there appeared to be an inverse relationship in some states, where a smaller population correlated with more funding per person. States with smaller populations, such as Alaska and North Dakota, received some of the highest funding per capita. This implied that funding allocation may be compensating for factors such as geographical size and infrastructure requirements that were not directly proportional to population size. Highly populous states, such as California and Texas, which stand out due to their large populations, did not rank first for funding per capita. This could imply that, while these states received a significant total allocation due to their size, the per capita distribution was lower than in less populous states, indicating an attempt to distribute funds more evenly across the country.

```{r state_funding, fig.height = 6}
sample_states <- first_df %>%
  arrange(desc(funding_per_capita)) %>% 
  #slice_head(n = 20) %>% 
  slice_head(n = 57) %>% 
  pull(abbriv)

sample_data <- first_df %>% 
  filter(abbriv %in% sample_states)

  
a <- ggplot(sample_data) + geom_col(aes(x=funding_per_capita, y=reorder(abbriv, funding_per_capita), fill=funding_per_capita)) +
  scale_fill_gradient(low="blue", high="red")+
  labs(title="Funding Per Capita by State/Territory", 
       x="Funding Per Capita, $", 
       y="State",
       fill="Funding Per Capita") + 
  theme_minimal() + 
  theme(axis.text.y  = element_text(size=6), legend.position="bottom", legend.text = element_text(size=6)) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_vline(xintercept=0, linetype="dashed")
```

```{r state_population, fig.height = 6}
sample_population <- first_df %>%
  arrange(desc(total_population)) %>% 
 # slice_head(n = 20) %>% 
  slice_head(n = 57) %>% 
  pull(abbriv)

sample_data <- first_df %>% 
  filter(abbriv %in% sample_population)

b <- ggplot(sample_data) +
  geom_col(aes(x=total_population, y=reorder(abbriv, total_population), fill=total_population)) +
  scale_fill_gradient(low="blue", high="red")+
  labs(title="Population by State/Territory", 
       x="Population", 
       y="State",
       fill="Population") + 
  theme_minimal() +
  theme(axis.text.y  = element_text(size=6), legend.position="bottom", legend.text = element_text(size=6)) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_vline(xintercept=0, linetype="dashed")

ggarrange(a, b, 
          ncol = 2)
```


### 2.2 Question 2: Political Bias in Funding

This scatter plot depicted the relationship between total funding allocation and population size while also reflecting the political landscape as determined by the 2020 Presidential Election results. Points were colored to indicate whether a state was won by Biden (blue) or Trump (red). Notably, the plot did not reveal a clear pattern in which one political party's winning states consistently receive more or less funding, implying that political victory did not always translate into financial favoritism. The states with the largest populations did not always receive the most funding, and vice versa, which may indicate that factors other than population and political leanings influenced funding distribution.

The bar plot revisited the average funding per capita by political affiliation, with a dashed line representing the overall average funding per capita across all states. It allowed for a direct comparison of state political affiliations (Biden vs. Trump) against the backdrop of the overall funding average. The inclusion of the overall average allowed us to assess whether the funding received by states won by Biden or Trump deviates significantly from the national norm. This graph showed that, while there could be differences in average funding per capita by political affiliation, both averages were close to the overall funding line, indicating a more nuanced distribution than a simple bias towards one political group.

```{r scatter_plot}
ggplot(second_df, aes(x=total_population, y=total_billions, color=factor(win))) +
  geom_point(aes(shape = factor(win)), size=4, alpha=0.7) +
  scale_color_manual(values=c('Trump'='red', 'Biden'='blue')) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title='Total funding vs. Population Size by 2020 Presidential Election Results',
       x='Population Size',
       y='Total funding, billions',
       color='Win',
       shape='Win') +
  theme_minimal() +
  theme(legend.position="right") +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_vline(xintercept=0, linetype="dashed")
```


```{r average_fund}
election_outcome <- second_df %>%
  group_by(win) %>%
  summarise(funding_per_capita = mean(funding_per_capita)) %>%
  as.data.frame()

avg_funding <- mean(second_df$funding_per_capita)

ggplot(election_outcome, aes(x=factor(win, labels=c('Trump', 'Biden')), y=funding_per_capita, fill=factor(win))) +
  geom_bar(stat="identity", position="dodge") +
   geom_hline(yintercept=avg_funding, color='darkviolet', linetype='dashed', linewidth=1) + 
  scale_fill_manual(values=c('blue', 'red')) +
  labs(title='Average Funding per Capita by 2020 Presidential Election Outcome',
       x='2020 Election Winner',
       y='Average Funding per Capita, $') +
  theme_minimal() +
  theme(legend.position="none") +
  scale_x_discrete(labels=c('Biden Wins', 'Trump Wins')) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_vline(xintercept=0, linetype="dashed")

```

## 3. Conclusions

While the allocation of Infrastructure Investment and Jobs Act funding was not strictly proportional to population size, it did appear to take into account factors other than sheer numbers. This approach may seek to address the unique needs and challenges of each state or territory, implying a multifaceted allocation strategy rather than a purely population-based one.

The analysis of funding allocation in relation to state populations and political outcomes from the 2020 Presidential Election revealed no clear, consistent pattern of political bias in fund distribution. Both visuals indicated that states won by either party received funding without a significant partisan skew. These findings indicate that, while politics will undoubtedly play a role in federal funding decisions, other non-partisan factors such as specific state needs, the cost of infrastructure projects, and regional economic considerations were likely to have a significant impact on how funds were allocated. To ensure transparency and fairness in federal financial distributions, these allocations must be monitored and analyzed on a continuous basis.

As we examine these allocations, the visual narratives highlight the importance of a data-driven approach to policy evaluation, recognizing that infrastructure needs vary across the country. Our findings suggest that while the allocation process may strive for equitable development, it also points to the need for transparency and ongoing analysis to fully understand the underlying principles guiding these significant federal investments.



## References

1) Knaflic, C. N. (2015). Storytelling with Data. https://doi.org/10.1002/9781119055259

2) US Census Bureau. (2023, December 18). State Population Totals and Components of Change: 2020-2023. Census.gov. https://www.census.gov/data/datasets/time-series/demo/popest/2020s-state-total.html

3) 2020 | The American Presidency Project. (n.d.). https://www.presidency.ucsb.edu/statistics/elections/2020

4) Cook Political Report. (n.d.). 2020 Popular Vote Tracker | Cook Political Report. https://www.cookpolitical.com/2020-national-popular-vote-tracker

5) Wikipedia contributors. (2024, February 1). List of U.S. states and territories by population. Wikipedia. https://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_population


6) Wikipedia contributors. (2024a, January 27). Native Americans in the United States. Wikipedia. https://en.wikipedia.org/wiki/Native_Americans_in_the_United_States


7) Ex-Pr. (n.d.). GitHub - ex-pr/DATA_608. GitHub. https://github.com/ex-pr/DATA_608/tree/main

8) The Data Visualisation catalogue. (n.d.). https://datavizcatalogue.com/index.html







